---
title: "What Drives Household Food Spending? A Regression Analysis"
author: "Parnika Raina"
date: "2026-01-19"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

## About This Project

An individual applied regression project using real household expenditure
data, working from a business question through data cleaning, modelling,
and interpretation to reach a conclusion.

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(emmeans) 
library(gridExtra) 
library(readr)

options(width=100)
```

### Business Problem

Understanding what drives a household's weekly food spending is useful for
anyone trying to model or predict household budgets. This analysis looks
at household characteristics and spending patterns, and specifically asks
whether the relationship between alcohol spending and food spending
changes depending on how much debt a household carries. If it does, debt
level would be an important segmentation factor; if it doesn't, one model
can be used across debt levels without splitting the population up.

### Data Dictionary

This data covers weekly household expenditure in England. The variables
are described below.

Variable | Description
------------- | -------------
Kids | Number of kids in the household
HouseholdSize | Number of members living in the household
Sex | The sex of the survey respondent (male/female)
Vehicles | Number of vehicles owned by the household
Food | Average amount spent on food per week
Housing | Average amount spent on accommodation per week
Alcohol | Average weekly spend on alcohol
Debt | Average weekly spend on debt repayments
income_category | The household's income level (Low/Medium/High)

### Data Overview

```{r}
#Reading the data
Household_data <- read_csv("HouseholdExpenditure_csv.csv")
```


```{r}
# Summary statistics for numerical variables
Household_data %>%
  select(where(is.numeric)) %>%
  summary()
```

```{r}
glimpse(Household_data)
```

```{r}
#Data integrity:-


missing_summary <- Household_data %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count")

print(missing_summary)


cat("Number of duplicate rows:", sum(duplicated(Household_data)), "\n")

```

A few variables have a handful of missing values, there are no duplicate
households, and the figures all look realistic - so filtering out the
missing values is the only cleaning needed before moving on.

### Filtering the data

```{r}

cat("Rows before filtering:", nrow(Household_data), "\n\n")
```
```{r}
# Removing the missing data in key variables
household_filtered <- Household_data %>%
  filter(!is.na(Food),           
         !is.na(Alcohol),        
         !is.na(Debt),
         !is.na(Vehicles),
         !is.na(Kids), 
         !is.na(Housing),
         !is.na(HouseholdSize), 
         !is.na(Sex),
         !is.na(income_category)) 
```

```{r}
  
# Converting categorical variables to factors

household_filtered <- household_filtered %>%
  mutate(
    Sex = factor(Sex),
    income_category = factor(income_category, 
                            levels = c("Low", "Medium", "High"), 
                            ordered = FALSE)
  )

cat("Rows after filtering:", nrow(household_filtered), "\n")
cat("Percentage of data retained:", 
    round(100 * nrow(household_filtered) / nrow(Household_data), 1), "%\n\n")


summary(household_filtered)
```
Almost every household was retained after filtering, so very little
information was lost and the data is still representative. Categorical
variables were converted to factors before modelling, so R treats them as
groups rather than as numbers.

### Demonstrating the Distribution of the Relevant Variables

### Food Distribution

```{r food_histograms, fig.width=12, fig.height=8, warning=FALSE}

  ggplot(household_filtered, aes(x = Food)) +
  geom_histogram(bins = 30, fill = "pink", colour = "black" ) +
  labs(title = "Distribution of Weekly Food Expenditure",
        x = "Food Expenditure (£/week)", y = "Number of Households") +
  theme_minimal()

```
Most households spend between £50 and £60 a week on food, and spending is
fairly balanced around that middle range rather than skewed by a few very
high or very low spenders. That makes a linear regression model a
reasonable choice here.

### Debt Distribution

```{r Debt_histograms, fig.width=12, fig.height=8, warning=FALSE}
  ggplot(household_filtered, aes(x = Debt)) +
  geom_histogram(bins = 30, fill = "lightblue", colour = "black") +
  labs( title = "Distribution of weekly debt payments", x = "Debt Payments (£/week)", y = "Number of Households") +
  theme_minimal()
```
Debt payments are less evenly spread - most households pay between £40 and
£70 a week, with fewer at the very low or very high end. Because debt
isn't evenly distributed, it makes sense to check whether households at
different debt levels show a different relationship between alcohol and
food spending.

### Alcohol Distribution

```{r Alcohol_histograms, fig.width=12, fig.height=8, warning=FALSE}
  ggplot(household_filtered, aes(x = Alcohol)) +
  geom_histogram(bins = 30, fill = "orange", colour = "black") +
  labs( title =  "Distribution of Weekly Alcohol Expenditure", x = "Alcohol Expenditure (£/week)", y = "Number of Households") +
  theme_minimal()
```
The average household spends between £15 and £20 a week on alcohol, with
relatively few spending a lot more or a lot less.

### Demonstrating the Relationships of the Relevant Variables

### Food vs Housing

```{r}
ggplot(household_filtered, aes(x = Housing, y = Food)) +
  geom_point(alpha = 0.4, color = "orange") +
  geom_smooth(method = "lm", color = "red") +
  labs( title= "Food Expenditure vs Housing Costs", x = "Housing (£/week)", y = "Food (£/week)")
```
Food spending tends to rise a little as housing costs rise, though the
relationship is fairly loose - useful as a control variable in a model,
but not a strong predictor of food spending on its own.

### Alcohol vs Food

```{r}
ggplot(household_filtered, aes(x = Alcohol, y = Food, colour = income_category)) + 
geom_point(alpha = 0.25) +
geom_smooth(method = "lm", se = FALSE) +
theme_minimal() +
labs(title = "Alcohol vs Food by Income category",
    x = "Weekly Alcohol spend (£)", y = "Weekly food spend (£)",
    colour = "Income category")

```
Across all income groups, food spending rises slightly as alcohol spending
rises, and the pattern looks much the same whether a household is on a
low, medium, or high income.

### Food vs Debt

```{r}
ggplot(household_filtered, aes(x = Debt, y = Food)) +
  geom_point(alpha = 0.4, color = "purple") +
  geom_smooth(method = "lm", color = "red") +
  labs( title = "Food vs Debt Payments", x = "Debt (£/week)", y = "Food (£/week)")
```

Debt on its own has only a weak relationship with food spending, which
suggests food spending isn't something debt alone can explain - a good
reason to build a model that accounts for several household factors at
once rather than looking at any one variable in isolation.

### Food vs Alcohol Across Debt Bands

```{r}
household_filtered %>%
  mutate(
    DebtBand = cut(
      Debt,
      breaks = quantile(Debt, probs = c(0, 0.33, 0.66, 1), na.rm = TRUE),
      labels = c("Low debt", "Medium debt", "High debt"),
      include.lowest = TRUE
    )
  ) %>%
  ggplot() +
  geom_point(
    aes(Alcohol, Food),
    alpha = 0.2,
    colour = "pink"
  ) +
  geom_smooth(
    aes(Alcohol, Food),
    method = "lm",
    se = FALSE,
    colour = "red"
  ) +
  facet_wrap(~ DebtBand) +
  labs(
    title = "Food vs Alcohol Across Debt Bands",
    x = "Weekly Alcohol Spend (£)",
    y = "Weekly Food Spend (£)"
  ) +
  theme_minimal()
```
Households at low, medium, and high debt levels all show a similar
pattern: food spending rises somewhat as alcohol spending rises. At a
glance, debt doesn't look like it's changing that relationship - worth
checking formally with a model.

### Model 1

```{r}
m.food <- lm(Food ~ HouseholdSize + Sex + Vehicles + Housing + Kids +
               Alcohol + Debt + income_category, data = household_filtered)

summary(m.food)
cbind(coef(m.food), confint(m.food))

```
This model looks at weekly food spending against household characteristics
and other spending, all together. It's statistically significant overall,
meaning these variables jointly help explain food spending.

The model explains about 15% of the variation in food spending - a modest
amount, but reasonable for household survey data where a lot of what
drives someone's spending simply isn't captured here.

Housing and alcohol spending are the strongest predictors: households that
spend more on either tend to spend more on food too, while higher debt is
linked to slightly lower food spending. Household size, number of
children, sex, and income category don't show a significant effect once
spending patterns are accounted for, and vehicle ownership's relationship
is weak and unclear.

### Model 2

```{r food_model_interaction}
m.interaction <- lm(
  Food ~ HouseholdSize + Kids + Vehicles + Housing +
    income_category + Sex + Alcohol * Debt,
  data = household_filtered
)

summary(m.interaction)
confint(m.interaction)
```
This model tests directly whether the relationship between alcohol and
food spending changes at different debt levels, by adding an
alcohol-by-debt interaction term. It's statistically significant overall.

Alcohol spending still has a positive relationship with food spending, and
housing remains the strongest predictor. But the interaction between
alcohol and debt itself isn't statistically significant - meaning debt
doesn't meaningfully change how alcohol relates to food spending. As
before, household size, number of children, sex, and income category
don't show a significant effect once spending is accounted for.

### Anova Test

```{r}
anova(m.food, m.interaction)
```
Adding the alcohol-debt interaction doesn't improve the model in a
statistically meaningful way. This confirms the simpler model without the
interaction is enough: alcohol's relationship with food spending doesn't
really depend on a household's debt level.

### Emmeans

```{r}
emmeans(
  m.interaction,
  specs = ~ Alcohol | Debt,
  at = list(
    Alcohol = c(10, 15, 20),
    Debt = quantile(household_filtered$Debt, c(.25,.50,.75))
  )
)

```
Predicted food spending rises as alcohol spending rises, at every debt
level. The size of that increase is similar whether debt is low, medium,
or high - consistent with the ANOVA result above, confirming debt doesn't
meaningfully change the alcohol-food relationship, even though households
with higher debt do tend to spend a little less on food overall.

### Conclusion

Food spending rises as alcohol spending rises, and this relationship holds
fairly consistently across low, medium, and high debt households - the
interaction between alcohol and debt was not statistically significant.
Housing and alcohol spending are the two strongest predictors of food
spending overall, while household size, number of children, sex, and
income category don't add much once spending patterns are accounted for.
For anyone modelling household food spending, this suggests debt level
doesn't need to be treated as a special segmentation factor for the
alcohol relationship specifically, though it's still a useful control
variable in its own right.

